\documentclass{cs4rep}
\usepackage{stmaryrd}


\begin{document}
\begin{thebibliography}{20}
\bibitem{kn:Plume} David Plume. {\em A Calculator for Exact Real Number Computation}. BSc Project. Edinburgh University. Availible from http://www.dcs.ed.ac.uk/~dbp/. 1998.
\bibitem{kn:Escardo} Mart\' \i n Escard\'o. {\em Effective and sequential definition by cases on the reals via infinite signed-digit numerals.} In Electronic Notes in Theoretical Computer Science, volume 13. Available from URL http://www.elsevier.nl/locate/entcs/volume13.html. 1998.
\bibitem{kn:DiGiantonio} Pietro Di Giantonio. A Golden Ratio Notation for the Real Numbers. CWI Technical Report. Availible from http://www.dimi.uniud.it/~pietro/Papers/paper\_arg.html.
\bibitem{kn:Longley} J.R. Longley. When is a functional program not a functional program?: a walkthrough introduction to the sequentially realizable functions. Standard ML source file, available from http://www.dcs.ed.ac.uk/home/jrl/, 1998.
\bibitem{kn:Longley2} J.R. Longley. When is a functional program not a functional program? Available from http://www.dcs.ed.ac.uk/home/jrl/, March 1999.
\bibitem{kn:DiGiantonio2} Pietro Di Giantonio. A Functional Approach to Computability on Real Numbers. Ph.D. Thesis: Universita di Pisa-Genova-Udine. Availible from http://www.dimi.uniud.it/~pietro/Papers/paper\_arg.html. 
\bibitem{kn:Menissier-Morian} Val\'erie M\'enissier-Morian. Abitrary precision real arithmetic: design and algorithms. {\em J. Symbolic Computation,} 11, 1-000. 1996.
\bibitem{kn:Beeler} M.Beeler, R.Gosper, R.Schroppel. {\em Hakmem}. Massachusetts Institute of Technology AI Laboratory. 1972.
\bibitem{kn:Nielsen} A.Nielsen, P.Kornerup. MSB-First Digit Serial Arithmetic. {\em Journal of Universal Computer Science}, Vol 1, no 7, 527-547. 1995.
\bibitem{kn:Boehm et al} H.Boehm. R.Cartwright. Exact Real Arithmetic, Formulating Real Numbers as Functions. In Turner. D., editor, Research Topics in Functional Programming, pages 43-64. Addison-Wesley, 1990.
\bibitem{kn:Boehm} H.Boehm. Constructive Real Interpretation of Numerical Programs. SIGPLAN Notices, 22(7):214-221, 1987.
\bibitem{kn:Simpson} A. Simpson. Lazy Functional Algorithms for Exact Real Numbers. {\em Mathematical Foundations of Computer Science 1998}, Springer LNCS 1450, pp. 456-464, 1998.
\bibitem{kn:Wiedmer} E.Wiedmer. Computing with Infinate Objects. {\em Theoretical Computer Science 10:133-155}. 1980.
\bibitem{kn:Potts} P.J. Potts. Computable Real Arithmatic Using Linear Fractional Transformations. {\em Imperial College of Science.} Available from http://theory.doc.ic.ac.uk/~pjp/. June 1996.
\bibitem{kn:Edalat} A.Edalat, P.J. Potts. A New Representation for Exact Real Numbers. In Electronic Notes in Theoretical Computer Science, volume 6. {\em http://www.elsevier.nl/locate/entcs/volume6.html.} 1997.
\bibitem{kn:Sunderhauf} P.S\"{u}nderhauf. Incremental Addition in Exact Real Arithmatic. In R. Harmer et al., editors, Advances in Theory and Formal Methods of Computing: Proceedings of the fourth Imperial College Workshop, IC Press. Availible from http://theory.doc.ic.ac.uk/~ps15/abstracts/incr.html. December 1997.
\bibitem{kn:Potts2} P.J. Potts, A. Edalat. Exact Real Computer Arithmatic. {\em Department of Computing, Imperial College.} Available from http://theory.doc.ic.ac.uk/~pjp/. March 1997.
\bibitem{kn:Graham} Graham, Knuth, Patashnik. Concrete Mathematics (A foundation for Computer Science). Section 6.6 (Fibonacci Numbers). {\em Addison Wesley.} ISBN 0-201-14236-8.
\bibitem{kn:Hall} J. Hall. Numerical Analysis and Optimisation. Course Notes. {\em Department of Mathematics, Edinburgh University.} 1998.
\bibitem{kn:bloke} Algorithms and Data Structures. Course Notes. {\em Department of Mathematics, Edinburgh University.} 1997.
\end{thebibliography}
\end{document}